Predicting Bike Usage at UofT St. George with Bayesian Poisson Regression

### 2024-12-14 paper/paper.qmd
Final checks
Final checks

There are many similar datasets available, but they pose significant limitations in the context of this study. Bike-sharing service providers worldwide have made their operational data publicly available for research purposes. For instance, the Shenzhen Municipal Transport Bureau in China released a dataset containing partial bike-sharing order information from January to August 2021. This dataset includes eight variables: user ID, start time, start longitude, start latitude, end time, end longitude, end latitude, and company ID, with a total of approximately 240 million records [@szdd]. While this dataset offers a rich variety of variables, it has limitations that make it unsuitable for this research. The data only covers an eight-month period, which is insufficient to capture long-term trends in bike-sharing usage. Additionally, university campuses in China are typically closed environments, unlike the open urban settings of Toronto, making the data less generalizable for this study's research context. Similar issues are present in other bike-sharing datasets from different regions. These include limited temporal coverage, which restricts long-term analysis, strong regional characteristics influenced by local cultural habits, and significant variations in data quality and structure due to fierce market competition among bike-sharing companies. These factors make it challenging to integrate data across companies or regions. Therefore, while these datasets may hold value for other studies, they do not meet the requirements of this research.

@@

@fig-daily_usage shows the daily usage totals of 27 shared bicycle stations from January 1, 2017, to September 30, 2024. It can be observed that the usage of the 27 bike-sharing stations at the University of Toronto's St. George campus has shown a steady upward trend, with an average increase of 1 use every 20 days. At the same time, there are clear seasonal variations: usage is higher in the spring and summer months (April to September) and significantly lower in the fall and winter months (October to March). Additionally, there was an abnormal surge in usage from June to September 2024, which can be attributed to two main reasons. First, historical data indicates that usage typically spikes in August and September each year, likely due to increased demand from students returning to campus for the start of the academic year. Second, during the same period in 2024, Bike Share Toronto added 52 new stations, most of which were located near the St. George campus, significantly enhancing the network's coverage and convenience[@bike]. These two factors combined led to the appearance of anomalies in August and September 2024.
```{r, fig.pos="H"}
#| label: fig-daily_usage
#| fig-cap: "Daily usage from January 2017 to September 2024"
#| echo: false
#| message: false
#| warning: false

@@

ggplot(merged_data, aes(x = station_name, y = count_difference)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    x = "Station Name",
    y = "Parking volume minus departure volume"
  ) +
  ylim(-4, 3) +
  coord_flip()
```

@@

This study considered using the ARIMA (autoregressive integrated moving average) model as an alternative but ultimately decided against it due to its limitations in capturing the specific characteristics of bike-sharing usage data. Although the ARIMA model has advantages in time series analysis, allowing for more accurate short-term predictions with parameters that have clearer interpretability. Bike-sharing usage exhibits significant seasonal patterns, time-of-day effects, and count-based characteristics, while ARIMA models are primarily suited for stationary time series with linear autocorrelations. ARIMA struggles to effectively address non-stationary trends and overdispersion, such as the marked decrease in usage during winter. Although extensions like SARIMA (Seasonal ARIMA) can account for seasonality, they still cannot directly handle the count-based nature of the data. In contrast, Bayesian Poisson regression is specifically designed for count data, allowing for the flexible integration of variables such as year, month, and specific four-hour intervals. Additionally, the Bayesian framework provides posterior distributions, offering deeper 
ideas into parameter uncertainty and variability, which are 
important for policy decisions. Therefore, Bayesian Poisson regression is more suitable for this study than the ARIMA model.

@@

In these environments, bike-sharing systems can be optimized to align with peak demand periods, such as shift changes in factories, scheduled breaks in office parks, or recreational times in residential areas. By leveraging predictive analytics administrators can strategically allocate bikes and docking stations to match demand patterns, ensuring users have access when and where they need it most.

@@

\appendix 
# Appendix {-}

@@

This study proposes an ideal data collection approach for bike-sharing systems, emphasizing key aspects such as data diversity, reliable data collection, anomaly detection and correction, and ethical considerations. Additionally, the study presents an ideal budget structure, a comprehensive checklist for data collection, and an ideal data framework.

@@

5. **Third-Party Services**
   - **API Integration:** Subscriptions for weather, traffic congestion, and air quality data APIs typically cost 1,000-5,000 CAD annually. Comprehensive API integration costs can range from 5,000-20,000 CAD annually.
   - **Privacy and Compliance Audits:** Engaging third-party firms to conduct privacy protection and compliance audits costs approximately 10,000-20,000 CAD per audit, with an annual budget of 20,000-50,000 CAD.

>>>>

There are many similar datasets available, but they pose significant limitations in the context of this study. Bike-sharing service providers worldwide have made their operational data publicly available for research purposes. For instance, the Shenzhen Municipal Transport Bureau in China released a dataset containing partial bike-sharing order information from January to August 2021. This dataset includes eight variables: user ID, start time, start longitude, start latitude, end time, end longitude, end latitude, and company ID, with a total of approximately 240 million records [@szdd].

While this dataset offers a rich variety of variables, it has limitations that make it unsuitable for this research. The data only covers an eight-month period, which is insufficient to capture long-term trends in bike-sharing usage. Additionally, university campuses in China are typically closed environments, unlike the open urban settings of Toronto, making the data less generalizable for this study's research context.

Similar issues are present in other bike-sharing datasets from different regions. These include limited temporal coverage, which restricts long-term analysis, strong regional characteristics influenced by local cultural habits, and significant variations in data quality and structure due to fierce market competition among bike-sharing companies. These factors make it challenging to integrate data across companies or regions. Therefore, while these datasets may hold value for other studies, they do not meet the requirements of this research.

@@

@fig-daily_usage shows the daily usage totals of 27 shared bicycle stations from January 1, 2017, to September 30, 2024. It can be observed that the usage of the 27 bike-sharing stations at the University of Toronto's St. George campus has shown a steady upward trend, with an average increase of 1 use every 20 days. At the same time, there are clear seasonal variations: usage is higher in the spring and summer months (April to September) and significantly lower in the fall and winter months (October to March).

Additionally, @fig-daily_usage shows an abnormal surge in usage from June to September 2024, which can be attributed to two main reasons. First, historical data indicates that usage typically spikes in August and September each year, likely due to increased demand from students returning to campus for the start of the academic year. Second, during the same period in 2024, Bike Share Toronto added 52 new stations, most of which were located near the St. George campus, significantly enhancing the network's coverage and convenience[@bike]. These two factors combined led to the appearance of anomalies in August and September 2024.

```{r, fig.pos="H"}
#| label: fig-daily_usage
#| fig-cap: "Daily usage from January 2017 to September 2024"
#| echo: false
#| message: false
#| warning: false

@@

ggplot(merged_data, aes(x = station_name, y = count_difference)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  labs(
    x = "Station Name",
    y = "Parking volume minus departure volume"
  ) +
  ylim(-4, 3) +
  coord_flip()
```

@@

This study considered using the ARIMA (autoregressive integrated moving average) model as an alternative but ultimately decided against it due to its limitations in capturing the specific characteristics of bike-sharing usage data. Although the ARIMA model has advantages in time series analysis, allowing for more accurate short-term predictions with parameters that have clearer interpretability, bike-sharing usage exhibits significant seasonal patterns, time-of-day effects, and count-based characteristics. ARIMA models are primarily suited for stationary time series with linear autocorrelations. ARIMA struggles to effectively address non-stationary trends and overdispersion, such as the marked decrease in usage during winter. Although extensions like SARIMA (Seasonal ARIMA) can account for seasonality, they still cannot directly handle the count-based nature of the data.

In contrast, Bayesian Poisson regression is specifically designed for count data, allowing for the flexible integration of variables such as year, month, and specific four-hour intervals. Additionally, the Bayesian framework provides posterior distributions, offering deeper thoughts into parameter uncertainty and variability, which are important for policy decisions. Therefore, Bayesian Poisson regression is more suitable for this study than the ARIMA model.

@@

In these environments, bike-sharing systems can be optimized to align with peak demand periods, such as shift changes in factories, scheduled breaks in office parks, or recreational times in residential areas. By utilizing predictive analytics administrators can strategically allocate bikes and docking stations to match demand patterns, ensuring users have access when and where they need it most.

@@

\appendix

# Appendix {-}

@@

This study proposes an ideal data collection approach for bike-sharing systems, emphasizing key aspects such as data diversity, reliable data collection, anomaly detection and correction, and ethical considerations. Additionally, the study presents an ideal budget structure, a thorough checklist for data collection, and an ideal data framework.

@@

5. **Third-Party Services**
   - **API Integration:** Subscriptions for weather, traffic congestion, and air quality data APIs typically cost 1,000-5,000 CAD annually. API integration costs can range from 5,000-20,000 CAD annually.
   - **Privacy and Compliance Audits:** Engaging third-party firms to conduct privacy protection and compliance audits costs approximately 10,000-20,000 CAD per audit, with an annual budget of 20,000-50,000 CAD.
